---
title: "VW_beta_diversity"
author: "ILF"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script compares and visualizes beta diversity metrics (Euclidean distances on CLR-transformed data: accounts for compositionality;UniFrac distances incorporates phylogenetic information) from a 16S rRNA dataset of gut microbiomes in from velvet worms, using a phyloseq object and plotting with ggplot2.

The goal is to assess whether and how the microbial communities differ by site.

```{r setup-2, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

##1 : Load Libraries
```{r}
library(vegan)
library(phyloseq)
library(cowplot)
library(viridis)
library(ggrepel)
library(ggplot2)
```

##2 : Load and Format Data

```{r load-data}
otuv = read.csv("silva.vw2.otu.csv", sep = ",", row.names = 1)
otu = as.matrix(otuv)

vtax = read.csv("silva.tax.csv", sep = ",", row.names = 1)
tax = as.matrix(vtax)

vmet = read.table("vwmeta2.csv", sep = ",", row.names = 1, header = TRUE)
tree = read_tree("silva.alldata.tree.nwk")

otu <- otu_table(otu, taxa_are_rows = TRUE)
tax <- tax_table(tax)
vfmet <- sample_data(vmet)
```

OTU table (silva.vw.otu.csv): contains counts of ASVs per sample.

Taxonomy (silva.tax.csv): assigns taxonomic identity to each OTU.

Metadata (vwmeta2.csv): sample-level metadata, including a "Site" field.

Tree: phylogenetic relationships between taxa.



##: 3 : Build phyloseq object and prune data
```{r load-data-2}
Q = phyloseq(otu, tax, vfmet, tree)
Q <- prune_taxa(taxa_sums(Q) > 0, Q)
```

Wraps the data into a unified object Q for downstream analysis.

Removes OTUs not present in any sample (zero counts across all samples).

## 4 : Compute for Two Types of Beta Diversity

A. Euclidean on CLR-transformed data
CLR corrects for compositional bias in sequencing data

```{r}
q <- Q
q_clr <- microbiome::transform(q, "clr")
clr_dist_matrix1 <- phyloseq::distance(q_clr, method="euclidean")
ord_euc <- ordinate(q_clr, method="MDS", distance="euclidean")

```
B. UniFrac on relative abundance data
UniFrac considers both community structure and phylogeny

```{r}
q_rel_abund = transform_sample_counts(q, function(x) x / sum(x))
unifrac_dist_matrix <- phyloseq::distance(q_rel_abund, method="unifrac")
ord_unifrac <- ordinate(q_rel_abund, method="PCoA", distance="unifrac")

```

## 5 :Plot the Ordination Results

Two ordination plots:

p1: Euclidean-CLR

p2: UniFrac

Each point = an individual velvet worm gut sample
Color = Site
```{r}
p1 <- plot_ordination(q_clr, ord_euc, color="Site") + 
  geom_point(size=2) +
  theme_minimal() + 
  scale_fill_manual(values=viridis(4)) + 
  ggtitle("Sites at ASV level: Euclidean-CLR") + 
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right",
    legend.title = element_blank()
  ) 
# Generate UniFrac distance matrix
unifrac_dist_matrix <- phyloseq::distance(q_rel_abund, method="unifrac")

# Ordination for UniFrac
ord_unifrac <- ordinate(q_rel_abund, method="PCoA", distance="unifrac")

# Plot ordination for UniFrac
p2 <- plot_ordination(q_rel_abund, ord_unifrac, color="Site") + 
  geom_point(size=2) +
  theme_minimal() + 
  scale_fill_manual(values=viridis(4)) + 
  ggtitle("Sites at ASV level: UniFrac") + 
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right",
    legend.title = element_blank()
  ) 
```


## 6: Combine the two plots
```{r}
combined_plot <- plot_grid(p1, p2, labels = "AUTO", ncol = 1)
combined_plot
```

## 7 Save the final plot 

```{r}

ggsave("Sites_Euclidean_CLR_and_UniFrac.pdf", combined_plot, width=7, height=9, units="in", dpi=300)


```


Summary: 

Key Concepts Involved
Beta diversity: Measures differences in microbial composition between samples

Ordination: Visual technique to simplify complex distance matrices (e.g., PCoA)

CLR: A compositional data transformation

UniFrac: Phylogenetic beta diversity metric (unweighted or weighted)

Main Insight You Get from This Script
Whether the gut microbial community composition of velvet worms is structured by site — and whether those differences are driven by relative abundances (CLR) or phylogenetic differences (UniFrac).
